---
title: "Installing R and R-Studio"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Posit-Cloud: R/R-Studio online

```{r child="zpg_T1_LaptopInstall.Rmd"}
```

------------------------------------------------------------------------------------------------------------------------

## Using R/R-Studio on your laptop

Optionally, you can use your own computer and laptop to complete the home-works and labs.

You need TWO pieces of software to run your labs.

-   The "R" Software teaches your computer to 'speak' R. This is why the R version is so important. Using an early
    version is like teaching your computer to speak Shakespearean English! The R-Studio software is like a programming
    version of Microsoft Word. It lets you work using the R language and has many nice features.

-   The "R_Studio" part is the program you actually use when completing your labs.

**QU. I already have R & R-Studio! Do I need to complete this section?**

-   YES! Both R and R-Studio change regularly, so I want everyone to be running the most up-to-date version and you will
    be graded on this in HW1. This should not hurt any of your other work. To update, follow the instructions below.

**QU. I have a very old computer, will this work?**

-   If you have a very old computer, this might not work. In that case, it's OK! You can use the Posit Cloud website.
    You're welcome to try the install and I'm happy to support you.

```{r child="zpg_T1_LaptopInstall.Rmd"}
```
