---
title: "R-Markdown reports"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

R Markdown is a powerful tool that allows you to combine text, code, and outputs in a single document. This tutorial will guide you through creating an R Markdown document, using templates, understanding its components, and working with key features like visual mode, code chunks, and knitting.

There are also many more hints and tips here <https://posit.co/blog/r-markdown-tips-tricks-1-rstudio-ide/>.


## Introduction to R-Markdown

```{r child="zpg_T7_Markdown_0Intro.Rmd"}
```

```{r child="zpg_T7_Markdown_1Structure.Rmd"}
```

## Creating an R-Markdown File

```{r child="zpg_T7_Markdown_2CreationTemplates.Rmd"}
```

## Editing YAML

```{r child="zpg_T7_Markdown_3EditYAML.Rmd"}
```

## Editing markdown text

```{r child="zpg_T7_Markdown_3EditText.Rmd"}
```

## Editing code

```{r child="zpg_T7_Markdown_4aCodeChunk.Rmd"}
```

```{r child="zpg_T7_Markdown_4bCodeChunkOptions.Rmd"}
```

## Knitting

```{r child="zpg_T7_Markdown_5Knitting.Rmd"}
```


