---
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    theme: flatly
---

```{=html}
<style>
h3 {
  margin-top: 15px !important;
  margin-bottom: 6px !important;
  padding-top: 0px !important;
  padding-bottom: 0px !important;
  line-height: 1.2 !important;
}
</style>
```

```{r, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(readxl)
library(kableExtra)
```

# Project 1 - PART 1. {.unnumbered}

## Introduction

Guardian Health is a new start-up company who is preparing to enter the US health insurance market in 2026. They need to
set competitive yet sustainable premiums for their plans and customers across the United States but because they are
new, they lack historical claims data. Instead, they ave contracted our consulting firm, *462-Investigations* to use the
Government’s MEPS longitudinal dataset (2022) to build a predictive model of total annual healthcare costs (TOTEXP22).

**As a junior analyst in STAT462-Investigations, your role on our team is to**

1.  Conduct the background research on the dataset and explore the data
2.  Use Simple Linear Regression to answer Guardian Health's questions about the expected annual medical expenses for
    their customers

**What will you submit?**

1.  You have already completed a quiz about Guardian health and the project (worth 20 points)

2.  You get 30 bonus points (previously the peer review) for participating in this roller coaster of a project

2.  By 23.59pm WEDNESDAY 23rd April, I expect you to submit TWO reports to submit to Guardian Health. (Worth 180 points):

    **Final Report (Client-Facing Document)\
    **This is a professionally formatted consulting report, ensuring that all results are interpreted and explained
    clearly for a non-statistical audience, including

    -   An introduction to the MEPS-2022 dataset, explaining its purpose, structure, and limitations.

    -   A summary of exploratory data analysis (EDA) and data quality checks.

    -   A full description of your Simple Linear Regression (SLR) models to answer the client questions, including
        rationale, key assumptions, and results.

    -   A conclusion discussing limitations and caveats, ensuring Guardian Health understands the scope and reliability
        of these predictions for the company's launch in 2026.. <br><br>\

    **Modelling Logs (Technical appendix/background workings)\
    **This is a record of your entire workflow and decision-making process, so I can see everything you tried, but you
    can keep the final report neat.\
    The logs can be written in a more casual style (including non English languages if English is your second language)
    and is designed to include things like

    -   All your messy quality control and exploratory data analysis

    -   Trying a load of models, then choosing your favorite for the main report

    -   All your checking/double checking assumptions, fixing transformations etc

<br>

[**All the instructions to create these reports are included below, and there is a detailed rubric at the
end.**]{.underline}

<br>

## Task 1. Create professional documents {.unnumbered}

THIS WOULD HAVE BEEN DONE AS PART OF YOUR WEEK 1&2 WORK

### 1.1 Create your project {.unnumbered}

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
First, create a project to store your data, analysis and reports. Read instructions on how to do this here

-   <https://psu-spatial.github.io/Stat462-2025/CH2_EACHLAB.html#1_R-Projects>
:::

### 1.2 Create your Lab reports {.unnumbered}

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
**A. Create your main report**

Create an Rmarkdown file for your final report and save to your project folder - NEW tutorial here:
<https://psu-spatial.github.io/Stat462-2025/CH2_EACHLAB.html#33_Creating_R-Markdown_Files>

You can be as detailed as you like and use any template, but at minimum\
- It should be clear that the report is written by you as part of STAT-462 Investigations - It should have a table of
contents - Unneeded output like library loading text printouts should be suppressed. - It should be saved as a filename
including your email id e.g. hlg5155 and the word REPORT

This means it is sufficient for you to copy/paste an old lab template if you wish. But bonus marks for making something
unique that you would be happy submitting to Guardian health in real life. Remember you have logo files etc if you wish
to use them, or images online (if you cite their source)

**B. Create your "background workings"/technical log report**

As I described above, this is where you can do all the 'messy' R analysis e.g. checking datasets, trying a load of
models.. etc. At minimum

-   It should be clear that the report is written by you as part of STAT-462 Investigations
-   It should have a table of contents
-   Unneeded output like library loading text printouts should be suppressed.
-   It should be saved as a filename including your email id e.g. hlg5155 and the word LOG

You can either use the method above, or simply copy/paste/rename your main report.
:::

<br>

### 1.3 Add code-chunk options {.unnumbered}

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
It's useful to stop warnings and messages being displayed. Try putting a code chunk looking like this at the top of BOTH
of your reports (below the YAML code at the top).

```{r, results='asis', echo=FALSE}
writeLines(c(
  "````markdown",
  "```{r, include=FALSE}",
  " knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)",
  "```",
  "````"
))
```
:::

<br>

### 1.4 Add libraries to both reports {.unnumbered}

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
You will need to load all the apps/libraries/packages you need (all different names for the same thing). I suggest
putting a code chunk, below your YAML code but at the top of your script, containing all the libraries you need. Here
are a few I suggest along with the associated code chunk options, but please add more as needed.

```{r, results='asis', echo=FALSE}
writeLines(c(
  "````markdown",
  "```{r, include=FALSE, message=FALSE, warning=FALSE}",
  "#-------------------------------------------------------------",
  "# Libraries",
  "#-------------------------------------------------------------",
  " library(tidyverse)",
  " library(remotes)",
  " library(ggplot2)",
  " library(ggstatsplot)",
  " library(dplyr)",
  " library(writexl)",
  " library(olsrr)",
  "```",
  "````"
))
```

If you don't have all these libraries on your computer, you might need to install them using the App-store accessed via
the Package-Tab "Install" button.
:::

<br>

### 1.5 Checklist {.unnumbered}

By now, you should have

1.  Made your project, and downloaded the data from canvas into the project folder - along with any logos etc as you
    wish.
2.  Created a nicely formatted main report file.
3.  Created a nicely formatting background workings log file.
4.  Added common libraries and code chunk options to both reports.

<br> <br>

## Task 2. Learn about the study {.unnumbered}

THIS WOULD HAVE BEEN DONE AS PART OF YOUR WEEK 1&2 WORK

First, you need to learn more about our client and their needs. Read the text below and note what they want to
achieve,their population of interest and any sub-populations they might care about.

### 2.1 Learn about Guardian Health {.unnumbered}

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
**Who are Guardian Health?**

Guardian Health is a (fictional) new company preparing to enter the US health insurance market in 2026. They aim to
redefine health insurance by offering transparent, data-driven, and customer-focused coverage. Their goal is to bridge
the gap between affordability and quality care, ensuring that individuals and families receive the support they need
without unnecessary financial burdens.

Guardian Health are particularly focused on reaching younger demographics, such as university students, and communities
that have traditionally faced barriers to affordable healthcare. To better serve these population, Guardian Health is
also hoping to develop specialized policies tailored to their unique needs.

**Innovation & Data-Driven Approach**

Guardian Health distinguishes itself by integrating machine learning and actuarial science to personalize insurance
plans. Their proprietary algorithms analyze consumer health trends, regional cost variations, and economic factors to
create fair and adaptable pricing structures. Additionally, they emphasize preventative care incentives, rewarding
policyholders who engage in wellness programs and routine health screenings.

Given the sensitivity of healthcare data, Guardian Health prioritizes ethical data use and privacy. They are committed
to maintaining rigorous security standards and complying with HIPAA and other industry regulations. They are aiming to
create their initial policies using the US Government MEPS dataset.

**Future Goals**

As a startup, Guardian Health need to set competitive yet sustainable premiums for their plans and customers across the
United States while expanding their reach into underserved communities. By 2026, Guardian Health aims to establish
itself as a reliable alternative to traditional insurers. Their roadmap includes:

-   Launching pilot programs in select states to test model effectiveness
-   Partnering with universities, community health organizations, and healthcare providers to expand coverage to
    students and low-income populations.
-   Expanding coverage options based on real-world data feedback.
-   Advocating for more transparent pricing in the insurance industry.

As Guardian Health moves closer to its market debut, they continue to refine their models and develop consumer education
initiatives to empower individuals in making informed healthcare choices.
:::

### 2.2 Learn about MEPS {.unnumbered}

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
**Introduction to MEPS**

The Medical Expenditure Panel Survey (MEPS) is a national US survey on healthcare utilization, expenditures, sources of
payment, and insurance coverage for the U.S. population.

The MEPS includes multiple components, with the **Full-Year Consolidated (FYC) file** being the primary dataset for
person-level analysis. We will be running some experiments using this file from data collected in 2022, which we can
access using R.

-   **Object**: An individual person
-   **Dataset/sample extent:** People treated in 2022? Randomly surveyed on the street? You need to find out!
-   **Sample Size:** Approximately **30,000 individuals**
-   **Total Variables:** Over **1,500** person-level variables
-   **Population Represented:** All U.S. civilian non-institutionalized population

<br>

**More reading**

-   MEPS Homepage: <https://meps.ahrq.gov/mepsweb/>

-   Read about MEPS here - <https://www.meps.ahrq.gov/mepsweb/about_meps/survey_back.jsp>

-   About the data: <https://meps.ahrq.gov/data_stats/data_overview.jsp>

-   Data Collection Process: <https://meps.ahrq.gov/survey_comp/hc_data_collection.jsp>

-   Importance and Uses:
    <https://odphp.health.gov/healthypeople/objectives-and-data/data-sources-and-methods/data-sources/medical-expenditure-panel-meps>

<br>
:::

### 2.3 Checklist {.unnumbered}

By now, you should

-   Understand more about our fictional client
-   Understand more about MEPS,ready to write up in your report (task 4)

<br> <br>

## Task 3. Access your dataset

THIS WOULD HAVE BEEN DONE AS PART OF YOUR WEEK 2 WORK

Originally we had code to read in MEPS from scratch. It turns out that data had many complexities, so your senior
analyst, Dr Greatrex, read in the data, did most of the quality control and saved it as an Excel file for you.

### 3.1 Read in the MEPS data

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}

We can use the `readxl` package to read in the dataset. I would start by putting this in my log file, then when I'm
happy, copying the code over to the main report.

1.  Download the excel file from Canvas - <https://psu.instructure.com/courses/2381686/assignments/17144942?module_item_id=44663028> . PUT IT IN YOUR PROJECT FOLDER <br>

2.  Make sure the **`readxl`** package is installed, that you have in your library code chunk (and that you have run
    it!) <br>

3.  Create a new code chunk and use the `read_excel()` function to read in the excel file and save it as a variable
    called `health22`. Tutorial here:
    <https://psu-spatial.github.io/Stat462-2025/in_T07_ReadingInData.html#14_Reading_in_excel_files> <br>

4.  Once you have done that, `health22` should appear in your environment. You can check it has loaded by running
    `summary(health22)` or by clicking on its NAME in the Environment tab. <br>

5.  Take a quick look at the summary to familiarise yourself with the data. You don't need to write anything yet.\
:::




<br>

### 3.2 Details about the data

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
There are over 1500 variables in MEPS! That's far too much for a class project, so here is a subset, chosen for their
general applicability and relevance to healthcare expenditures and insurance coverage. You can read more generally about
MEPS here.

**YOUR RESPONSE VARIABLE:**

-   **TOTEXP22**: Total healthcare expenditures in 2022 (USD)
    -   <https://meps.ipums.org/meps-action/variables/TOTEXP22>

**POTENTIAL PREDICTOR VARIABLES:**

-   **PID**: Person ID. Unique identifier for each individual in the dataset. Or is it??
    -   <https://meps.ipums.org/meps-action/variables/PID>
-   **AGELAST**: Age at the end of 2022
    -   <https://meps.ipums.org/meps-action/variables/AGELAST>
-   **SEX**: Gender (1 = Male, 2 = Female), no other options included in survey
    -   <https://meps.ipums.org/meps-action/variables/SEX>
-   **EDUCYR**: Years of education completed
    -   <https://meps.ipums.org/meps-action/variables/EDUCYR>
-   **FAMINC22**: Total family income in 2022 (USD)
    -   <https://meps.ipums.org/meps-action/variables/FAMINC22>
-   **ADBMI42**: Adult Body Mass Index (BMI), based on self-reported height and weight
    -   <https://meps.ipums.org/meps-action/variables/ADBMI42>
-   **OBTOTV22**: Total number of office-based medical provider visits in 2022
    -   <https://meps.ipums.org/meps-action/variables/OBTOTV22>
-   **VMCS42**: Mental component summary score (VR-12) – a standardized score of mental health status
    -   <https://meps.ipums.org/meps-action/variables/VMCS42>
-   **VPCS42**: Physical component summary score (VR-12) – a standardized score of physical health status
    -   <https://meps.ipums.org/meps-action/variables/VPCS42>
-   **UNINS22**: Uninsured status in 2022 (1 = uninsured all year, 2 = any coverage)
    -   <https://meps.ipums.org/meps-action/variables/UNINS22>

\*\* DATA QUALITY FLAGS\*\*

All the columns show similar data quality flags. See section 2.2 pg 13 here -
<https://meps.ipums.org/meps/resources/h233doc.pdf>

-   -1 Inapplicable (question not asked, due to skip pattern)
-   -2 Don't know
-   -3 Invalid missing (e.g., multiple answers or problem in data collection)
-   -7 Refused
-   -8 Not ascertained (unable to obtain)
-   -9 Not known – unspecified (used rarely)
-   -15 Suppressed due to confidentiality (e.g., too few cases in a cell)
:::

<br>

### 3.3 Finish the quality control

I did most of the quality control, but there are a few things left for you to do.

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
Again, I would work out how to do this in my log/background file, then when I'm happy, copy the code over to the main
report.

**A. In your background workings, Create a section called quality control**

<br>

**B. Missing Data**

For the sake of an undergrad project, I assumed that all quality control codes (see above) that were NOT -1 should be
treated as missing data, NA. This means that R will ignore those values.

Unfortunately I forgot to assign -8 to NA. This code will do this for you, so add this to your quality control section
and see how it changes the summary of the data.

```{r, eval=FALSE}


 special_code <- -8
 
 health22 <- health22 %>% 
               mutate(across(everything(), ~ replace(.x, .x %in% special_code, NA)))
 
```

<br>

**C. Categorical data and Factors**

Some of our data is grouped/categorical and we need to make sure R understands it as such.

Include this code to make your uninsured data and sex columns categorical:

```{r, eval=FALSE}
# because it already has labels
health22$UNINS22 <- as.factor(health22$UNINS22)

health22$SEX <- factor(health22$SEX, levels = c(1,2), 
                             labels=c("Male", "Female"))


# set this to a descriptor
health22$PID <- as.character(health22$PID)

```

<br>
:::

<br>

### 3.4 Exploratory data analysis

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
Now focus on exploring the data in your log file, using commands like `summary(health22)` and other statistics.

-   More commands here - <https://psu-spatial.github.io/Stat462-2025/CH3_EDA.html#ch8>
-   Filtering/selecting commands here: <https://psu-spatial.github.io/Stat462-2025/CH3_EDA.html#ch9>
-   Many plots here <https://psu-spatial.github.io/Stat462-2025/in_T11_Plots.html>

YOU DON'T NEED TO DO CORRELATIONS YET. (its fine if you have)

Record notes in your logfile about what you have found.
:::

### 3.4 Checklist {.unnumbered}

By now, you should

-   Have read your data into your log report
-   Finished some more data wrangling
-   Explore the data a little

<br><br>

## Task 4. Write report introduction {.unnumbered}

THIS WOULD HAVE BEEN DONE AS PART OF YOUR WEEK 2 WORK

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
In your main report, create a clearly labeled section called **“Introduction”** or **“Background”**.

**Underneath, copy over a 'tidy' version of your logbook code.**\
What do I mean by 'tidy'? You need to read in the dataset and do all your quality control, but you don't need to include
every single printout. You can also use code chunk options like output="hide" , include=FALSE or code folding to keep it
neat

Then, write an introduction for Guardian-Health, explaining in clear non-expert terms what they are trying to achieve.

This section should be around 300–500 words (min) and written in a clear, professional style that someone at Guardian
Health (your client) could understand and learn from. Please use formatting tools like headings/sub-headings/bullets etc
to make your write up look professional. You are also allowed to use any images you wish

Your goal is to give the reader a strong overview of the project before you dive into the data or analysis. Keep this
section focused on the big picture. Don’t include specific numbers, variables, or plots yet — that comes in the next
section where you explore the data.
:::

### 4.2 Checklist & rubric {.unnumbered}

**If you did ANYTHING Guardian Health or I might find interesting on this dataset, then please keep it in your reports
You don't have to delete any old analysis - and I regularly award bonus points for interesting things you have found out
about the data.**

Top introductions will:

1.  Clearly introduce the background for Guardian health. Identify the population Guardian Health are hoping to
    extrapolate their results to. (All humans who ever existed?)

2.  Describe MEPS, its background and how it's created - along with any relevant/interesting information.

3.  Describe the actual excel dataset e.g. the

    -   the object-of-analysis,
    -   the dataset extent
    -   and key variables (you can copy/paste my list from above)
    -   explain if it's a sample or a census and why.

4.  Note any initial limitations e.g. does MEPS generalise to all populations? Do you have issues with individual
    variables/subpopulations?

5.  Make sure the code runs in the main report to read in and quality control the data, then discuss any initial thoughts or potential limitations. Do you have issues with individual variables/subpopulations? Will our results help Guardian health with all their populations? If not, why not.

Here's how I am grading this section. Importantly, I am grading based on content, not on grammar/English skills.

```{r, echo=FALSE}
rubric <- readxl::read_excel("Table_ProjectRubric.xlsx")
knitr::kable(rubric) %>%   
  kable_classic_2() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))


```

<br><br>

## Task 5. Exploratory Data Analysis {.unnumbered}

THIS WOULD HAVE BEEN DONE AS PART OF YOUR WEEK 2 WORK


### 5.1 Analyse TOTEXP22 {.unnumbered}

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
The response variable that Guardian health care about is total health care expenses in 2022, the variable `TOTEXP22`.

In your log files, look carefully at this variable. I have also included the natural log transformed version of the
variable, `ln_TOTEXP22`. For example, look at the summary statistics in the context of what people spend, the histogram
and anything else.

Comment on what you find, how that results to real life and any issues you see regarding using this for Simple Linear
Regression. When you are happy, create a new section in your main report and transfer over your neat code and write up
your findings.

### 5.2 Remove missing values and non-numeric columns {.unnumbered}

For the sake of undergraduate project easiness, let's remove all missing values and non-numeric columns.

```{r,eval=FALSE}
# remove all missing data
  health22 <- na.omit(health22)

# remove all non-numeric data
  health22 <-  health22 %>% select(where(is.numeric))


```
:::

<br>

### 5.2 Look at correlations with TOTEXP22 {.unnumbered}

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
Now in your log file, explore the relationships between TOTEXP22 and different predictor variables. This code makes the core correlation matrix

```{r,eval=FALSE}
cortable <- cor(health22_numeric,use="pairwise.complete.obs")
cortable
```

-   You could use correlation matrices: <https://psu-spatial.github.io/Stat462-2025/in_T13_Correlation.html>
-   Or make individual scatterplots <https://psu-spatial.github.io/Stat462-2025/in_T11_Plots.html#4_Scatterplots> (only
    have to be neat in your final report)

If you want to see all the scatterplots in one place, you can use this code - where you change COLNAME to the name of
the column you wish to explore e.g. TOTEXP22.  Or you  could just copy/paste the scatterplot code or change the variable name to take a look at them.

```{r, eval=FALSE}

# THIS TAKES A MINUTE TO RUN!

# Rearrange the format
 health22_longformat <- health22 %>%
   pivot_longer(cols = -COLNAME,   #  KEEP THE -sign in front!  e.g. -TOTEXP22
                names_to = "predictor", 
                values_to = "value")
 
 # then plot using ggplot
 ggplot(health22_longformat, aes(x = value, y = COLNAME)) +
   geom_point(alpha = 0.3,size=1) +
   facet_wrap(~ predictor, scales = "free_x") +
   theme_minimal()

```
:::

<br>

### 5.3 Look at correlations with ln_TOTEXP22 {.unnumbered}

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
Now in your log file, explore the relationships between TOTEXP22 and different predictor variables.

-   You could use correlation matrices: <https://psu-spatial.github.io/Stat462-2025/in_T13_Correlation.html>
-   Or make individual scatterplots <https://psu-spatial.github.io/Stat462-2025/in_T11_Plots.html#4_Scatterplots> (only
    have to be neat in your final report)
-   Or use the code from 5.3 to see all the scatterplots in one place,
:::

<br>

### 5.4 Look at correlations with non-zero health costs {.unnumbered}

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
Finally, it appears that many people paid zero for their health costs (e.g. they were healthy??). This skews our
histogram and its reasonable to assume that the predictors causing someone to be yes/no unhealthy are different to the
predictors causing someone to pay more money if they are.

This code will filter the data to ONLY positive healthcare costs.

```{r,eval=FALSE}
  health22_TOTEXP22_gt0 <-  health22[which(health22$TOTEXP22> 0),]

```

get this code running and repeat your quality control. Remember you can copy/paste code.
:::

<br><br>


### 5.5 Write up what you have found in the main report {.unnumbered}


Create a new section in your main report called `Quality Control and EDA` (or similar). This is where you’ll present the final version of your exploratory data analysis and quality control steps. 

**Style**
 - Use clear subheadings, bullets, and brief comments so your client (Guardian Health) can follow along.
 - Only include plots/tables that help tell a clear story. You can always refer clients back to your log files.

#### Code to include: {.unnumbered}

 - Code to read in the dataset (if not already in your main report)
 - Code for any quality control/cleaning steps (e.g. removing missing data or non-numeric columns)
 - Code for summary statistics and distributions for TOTEXP22 and key variables
 - Code for correlations and/or scatterplots with TOTEXP22 and ln_TOTEXP22
 - Code for your exploration of cases with TOTEXP22 > 0

#### What to write:  {.unnumbered}

1. Explain what you did for quality control and why. 
2. Explain what you did for exploratory data analysis.
3. Highlight any key insights or limitations (e.g. missing data, skewed variables, zero costs, strange variables). 
4. Reflects on what this means for future modeling and Guardian Health's goals

You don’t need to include everything — focus on what’s useful and interesting.



:::

### 5.6 Checklist & rubric {.unnumbered}

Top write-ups will:

 - Clearly explain QC and EDA steps taken

 - Include relevant summaries or visualizations

 - Note any data issues or surprises

 - Use neat, well-organized code

 - Reflect on what this means for the analysis and the client

Here's how I am grading this section. Importantly, I am grading based on content, not on grammar/English skills.

```{r, eval=FALSE}
rubric <- readxl::read_excel("Table_ProjectRubric.xlsx")
knitr::kable(rubric) %>%   
  kable_classic_2() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))


```

<br><br>

## Task 6 SLR question 1 {.unnumbered}

### 6.1 Age and total expenses {.unnumbered}

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
Guardian health would like to know the impact of AGE (in 2022) on expenses paid. You can choose whatever variant of the
response variable you wish, but you must justify your choice.

-   You can find the code here for SLR <https://psu-spatial.github.io/Stat462-2025/in_T14_Regression.html>
-   You can find the code here for model diagnostics
    <https://psu-spatial.github.io/Stat462-2025/in_T15_Diagnostics.html>
-   You can find the code here for prediction <https://psu-spatial.github.io/Stat462-2025/in_T17_ConfPred.html>

HINT, I first suggest taking a look at AGELAST if you haven't already and putting your findings in your exploratory data analysis 

*Advanced - We talked for a long while in week 2 lab about some of the features of this data).  If you decide to filter the dataset, you can do it the same way we removed zeros above or see the tutorial on filtering and selecting.*



Specifically, please write up

1.  The causal chain - why would age lead to differing healthcare costs?
2.  *Advanced - Any quality control/filtering you performed on the AGELAST data, why and what it means for your study* 
2.  The model equation
3.  The slope and intercept in their real life meaning
4.  The effect size and significance of the slope.
5.  Whether the model meets the LINE assumptions and if not what you suggest (you dont have to do it)
6.  Any issues/outliers - what about very old people?
7.  Predict the 95% confidence interval on total healthcare costs for 40 year olds.
:::

<br>

### 6.2 "Best model" {.unnumbered}

If you want to compete.. here is the best R2, RMSE that I found a single predictor so far.



<button class="accordion">

Expand for Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}

Guardian health would like to know the MOST IMPACTFUL predictor (in 2022) on expenses paid. You can choose whatever
variant of the response variable you wish, but you must justify your choice.

-   You can find the code here for SLR <https://psu-spatial.github.io/Stat462-2025/in_T14_Regression.html>
-   You can find the code here for model diagnostics
    <https://psu-spatial.github.io/Stat462-2025/in_T15_Diagnostics.html>
-   You can find the code here for prediction <https://psu-spatial.github.io/Stat462-2025/in_T17_ConfPred.html>

Specifically, please write up why you chose that variable., Plus!

1.  The causal chain - why would your chosen predictor lead to differing healthcare costs?
2.  The model equation and why you chose it.
3.  The slope and intercept in their real life meaning
4.  The effect size and significance of the slope.
5.  Whether your model meets the LINE assumptions.
6.  Feel free to use transformations on either/both the response and predictor if needed
7.  Any issues/outliers - what about very old people?
8.  Use AIC and R2 to compare your age model with this one. (optional, predicted R2 from HW8)


:::

<br>

### 6.3 Write up your results {.unnumbered}

In your main report, create a clearly labeled section called “Simple Linear Regression” or “SLR Models”. Underneath, copy over your neat code and write-up for the two models.  Just like before, keep things tidy and only include plots and outputs that help tell your story.  

Optional, try playing with code chunk options to show/hide code.

For each model, write clearly and professionally so someone at Guardian Health could understand your choices and results.  

#### For AGE - your write-up should include:  {.unnumbered}

 - A brief causal explanation (why the predictor might affect total annual expenses)

 - The regression equation, written out using variable names your client will recognize, making it clear if this is the sample regression or population regression line - and which variant of the response variable you chose.

 - Real-life meaning of the slope and intercept

 - Effect size (the real life impact of the slope, see Lab 3) vs and statistical significance of the slope (e.g. confidence intervals/T-tests), 

 - A check of model assumptions (LINE) and whether they hold alongside any strategies if they don't.

 - A prediction (with confidence interval) for a specific example (e.g. a 40-year-old)
 
 <br>

#### For your best model - your write-up should include:  {.unnumbered}






<br><br>

## 7. Conclusions

Finally, write a short summary/conclusion in your main report.

If you could choose several predictors, state what you would choose and why




### BONUS questions for an A**

You don't need to answer these for an A, but you do for a top 100% score

 - Its often said that BMI (body mass index) will impact how healthy someone is. Do you see evidence to support this on total healthcare costs? 
 
 - Does mental health or physical health have more of an impact on total healthcare costs? 
 
 - What do you think the influence of COVID is on these results and Guardian health's ambitions?
 
 - 
 

# Grading rubric

<button class="accordion">

Expand For Instructions

</button>

::: {.panel style="display: none; margin-top: 10px;"}
Overall, here is what your lab should correspond to:

```{r, echo=FALSE}
rubric <- readxl::read_excel("Table_ProjectRubric.xlsx")
knitr::kable(rubric) %>%   
  kable_classic_2() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))


```
:::

```{=html}
<script>
document.addEventListener("DOMContentLoaded", function() {
  document.querySelectorAll('.accordion').forEach(button => {
    button.addEventListener('click', function() {
      const panel = this.nextElementSibling;
      if (panel.style.display === "none" || panel.style.display === "") {
        panel.style.display = "block";
      } else {
        panel.style.display = "none";
      }
    });
  });
});
</script>
```

```{=html}
<style>
.accordion {
  background-color: #f2f2f2;
  color: #333;
  cursor: pointer;
  padding: 10px;
  width: 100%;
  text-align: left;
  border: none;
  outline: none;
  transition: background-color 0.3s ease;
  font-size: 16px;
  margin-bottom: 2px;
  border-radius: 4px;
}

.accordion:hover {
  background-color: #ddd;
}

.panel {
  padding: 10px;
  background-color: #fafafa;
  border-left: 2px solid #ccc;
  border-radius: 4px;
}
</style>
```
