```{r setup, include=FALSE,message=FALSE,warning=FALSE}
# OPTIONS -----------------------------------------------
knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      message = FALSE)

# PACKAGES-----------------------------------------------
# Tutorial packages
library(vembedr)
library(skimr)
library(yarrr)
library(RColorBrewer)
library(GGally) 
library(tidyverse)
library(plotly)
library(readxl)
library(rvest)
library(biscale)
library(tidycensus)
library(cowplot)
library(units)
library(olsrr)

data("HousesNY", package = "Stat2Data")



```

# Regression 

Now we will fit our first regression model.

## "Standard" regression output 

The command to do this is `lm()` e.g. linear model.

```{r,eval=FALSE}
output <- lm(y_column ~ x_column,data=tablename)
output
```

NOTE, THE WEIRD \~ SYMBOL. This means "depends on"/"is modelled by" and it's how we tell R what the response variable is. E.g. in y=mx+c, y depends on x, or y is modelled/predicted by x.

For example for the NYHouses dataset, if I wanted to create a model of the Price and Lot columns to see if Lot size could predict house sales price, then I would type.

```{r}
# response = Price, predictor = Lot size
Model1.lm <- lm(Price ~ Lot,data=HousesNY)
Model1.lm
```

So we are saying here that the equation is

Expected_Average_Price = -0.5749\*Lot_Size + 114.0911

E.g. the average expected price house with no Lot/Garden is 114.09 (x1000 USD)


### Printing out the equation

Now you have the numbers, you can use Tutorial 6 on Markdown to print out the equation. For example, if you wanted to print out the equation of the line, you could use the following code:

$$
\hat{y} =  114.0911  -0.5749 \times \text{Lot}
$$



## Model summary

You can  look at the summary by using the `summary()` command:

```{r}
summary(Model1.lm)
```

<br>

### OLSRR regression summaries

If you want a different way of seeing the same output, you can use the `ols_regress()` command inside the `olsrr` package.  THIS ONLY WORKS IF YOU USED lm(y ~ x, data=TABLENAME).  If you used a $ it will break.

```{r}
library(olsrr)
Model1.lm.ols <- ols_regress(Model1.lm)
Model1.lm.ols
```

The ols_regress command produces beautiful output, but sometimes it doesn't work well with other commands. So I tend to run a lm command at the same time to have both available.


<br>

## Plotting your regression line

See the scatterplot tutorial under plots!

