---
output:
  html_document:
    toc: yes
    toc_depth: 2
    number_sections: yes  
    toc_float:
      collapsed: no  # Ensure Level 1 & 2 remain expanded
      smooth_scroll: yes
---

## All the commands

Before starting, ensure you have these packages installed and loaded:

```{r,warning=FALSE,message=FALSE}
library(skimr)   # For detailed data summaries
library(dplyr)   # For data manipulation
library(naniar)  # For missing data visualization
library(tidyr)   # Lots of useful functions for data cleaning
```

<br>

Here are the commands covered in this guide to make it easy to copy/paste. Remember to replace `testdata` with the name
of your dataset - and keep reading for what each command does.

```{r, collapse=TRUE, eval=FALSE}
#-------------------------------------------
# Visualizing Missing Data
#-------------------------------------------
summary(testdata)        # Summary statistics
skim(testdata)           # Skim summary (detailed)
colSums(is.na(testdata)) # Count NAs per column
rowSums(is.na(testdata)) # Count NAs per row
sum(is.na(testdata$x))   # Count NAs in column x
gg_miss_var(testdata)    # naniar: Missing values per variable
vis_miss(testdata)       # naniar: Heatmap of missing values

#-------------------------------------------
# Removing Rows with Missing Data
#-------------------------------------------
data_complete <- na.omit(testdata)
data_filtered <- testdata %>% dplyr::filter(!is.na(x) & !is.na(z)) 

# using tidyr
testdata <- testdata %>% replace_na(list(x = 0, y = "Unknown"))
testdata <- testdata %>% drop_na(x, y)  

#-------------------------------------------
# Replacing missing values. In this case
# Replace NAs with 99
#-------------------------------------------
testdata$x <- ifelse(is.na(testdata$x), 99, testdata$x)  


#-------------------------------------------
# Handling NAs in Common Functions
#-------------------------------------------
meanx <- mean(testdata$x, na.rm=TRUE)   
sdx <- sd(testdata$x, na.rm=TRUE)       
table(testdata$y, useNA = "ifany")       
cor(testdata, use = "pairwise.complete.obs")  
lm(z ~ x, data = testdata, na.action = na.omit)    # Omitting NAs
lm(z ~ x, data = testdata, na.action = na.exclude) # Keeping track of NAs, prefer

```

<br><br>

## What is Missing Data

Before making decisions about how to handle missing data, you need to identify where it's present. In R, `NA` represents
missing values. You might also encounter `NaN` (Not a Number), which is another type of missing value.

#### Many ways of writing 'missing' {.unnumbered}

When reading in external datasets, missing values may be represented differently. Common placeholders include:

-   `-1` (often used in government datasets)
-   `99`, `-99`, `9999`, `999` (arbitrary placeholders for missing values)
-   Blank cells in spreadsheets

### There are different types of missing

Understanding why data is missing is crucial before deciding how to handle it:

1.  **Missing Completely at Random (MCAR)**
    -   The probability of missingness is unrelated to any variable in the dataset.
    -   Example: A sensor randomly fails.
2.  **Missing at Random (MAR)**
    -   Missingness is related to observed data but not the missing values themselves.
    -   Example: Younger participants are less likely to report their income, but within each age group, missingness is
        random.
3.  **Missing Not at Random (MNAR)**
    -   Missingness depends on the value of the missing data itself.
    -   Example: People with higher incomes are less likely to report their income.

**Key Takeaways:** - If data is **MCAR**, deleting missing cases is often unbiased. - If data is **MAR**, multiple
imputation methods may help recover missing values. - If data is **MNAR**, the missingness mechanism itself needs to be
modeled.

<br><br>

## Files with missing data

If you know your dataset uses non-standard missing value indicators, you can explicitly define them when reading in the
data. No matter what, check the data carefully to make sure you didn' make a mistake.

#### CSV Files {.unnumbered}

```{r, eval=FALSE}
data <- read.csv("data.csv", na.strings = c("", "NA", "-1", "99", "9999"))
```

-   `na.strings` tells R which values should be treated as `NA`.

-   This example converts empty strings (`""`), standard `NA`, `-1`, `99`, and `9999` into missing values.

<br>

#### Excel files {.unnumbered}

```{r, eval=FALSE}
library(readxl)
data <- read_excel("data.xlsx", na = c("", "NA", "-1", "999"))
```

-   `na` defines missing values similarly to `na.strings` in `read.csv()`.

<br>

#### Reading Data with `tidyverse` (readr package) {.unnumbered}

```{r, eval=FALSE}
library(readr)
data <- read_csv("data.csv", na = c("", "NA", "-1", "999"))
```

-   `read_csv()` is part of `tidyverse` and often preferred for faster performance and better defaults.

<br><br>

## Summarising Missing Data

Before deciding how to handle missing data, it's crucial to first identify where it's present. Let's create a small
dataset with missing values. The final column includes many empty values to mimic real-world data.

```{r}
testdata <- data.frame(
  x = c(1, 2, 3, NA, 5),
  y = c("A", "B", "C", "D", NA),
  z = c(10, 15, NA, 20, 25),
  randomnotes = c(NA, NA, NA, "hello", NA)
)
print(testdata)
```

along with example outputs. I apply everything to the basic test data above.

<br>

#### General Summaries {.unnumbered}

Summary functions provide an overview of missing data:

```{r}
summary(testdata)  # Basic summary statistics
skim(testdata)     # More detailed summary, including missing values
```

<br>

#### Visualizing NA numbers (naniar package) {.unnumbered}

The `naniar` package provides helpful visualizations. Install and load it before use.

```{r}
gg_miss_var(testdata)  # Missing values per variable
vis_miss(testdata)     # Heatmap of missing values
```

<br>

#### Counting NAs {.unnumbered}

You can count missing values at different levels:

```{r}
colSums(is.na(testdata))            # Count NAs per column
rowSums(is.na(testdata))            # Count NAs per row
sum(is.na(testdata$x))              # Count NAs in a specific column

# Frequency table including missing values
table(testdata$y, useNA = "ifany")
```

<br><br>

## Handling Missing Data

Missing data can cause errors in analyses, so it is often necessary to handle it appropriately. Removing missing data is
only advisable when the data is **Missing Completely at Random (MCAR)**.

<br>

### 1. Remove entire row if NA present {.unnumbered}

**Listwise deletion**, or **Complete Case Analysis**, removes entire rows if any column contains missing data. This can
significantly reduce your dataset, so use it with caution.

```{r}
data_complete <- na.omit(testdata)  # Remove rows with any NA

# Check for potential bias introduced by removal:
summary(data_complete)
```

**Warning:** If many columns have missing data (especially ones you don't care about, such as "notes"), then this method
may remove too many observations. For example in this case, I removed most of my valuable information.

<br>

### 2. Removing Rows with NAs in Chosen Columns {.unnumbered}

If you want to only remove rows with missing values in a single column, we can use the complete cases command. E.g. take
the testdata table and ONLY include objects that are not missing their x-value. Note, this can be any column name at
all.

```{r}
data_filtered <- testdata[complete.cases(testdata$x), ]
```

You can also do this manually using `filter` and the `is.na()` command

```{r}
# Tidyverse
data_filtered <- testdata %>% 
  dplyr::filter(!is.na(x) & !is.na(z))

# BaseR
data_filtered <- testdata[which(!is.na(testdata$x) & !is.na(testdata$z)), ]
```

<br>

### 3. Turning values into NA

Sometimes, you have -999 or something as your missing datacode and you missed it reading in the data. You can also use
the filter command to force data to be NA.

E.g. here I select all the prices in a house dataset that are less than 0 (e.g impossible) and set them to NA.

```{r, eval=FALSE}
HousesNY$Price[HousesNY$Price < 0]   <- NA
```

E.g. here I select all the prices that are less than 0 (e.g impossible) and set them to NA.

<br>

### 3. Replace NAs with data {.unnumbered}

This is useful when you want to keep the row but replace the missing value with a specific number.

```{r}
testdata$x <- ifelse(is.na(testdata$x), 99, testdata$x)  # Replace NAs with 99
```

<br>

### 4. Use the `tidyr` package {.unnumbered}

The `tidyr` package provides a more elegant way to replace NAs. Remember to load/install it first!

**Using `tidyr::replace_na()`**

```{r}
testdata <- testdata %>% replace_na(list(x = 0, y = "Unknown"))
```

**Filtering Out NAs Using `drop_na()` (tidyr)**

```{r}
testdata <- testdata %>% drop_na(x, y)  # Removes rows where x or y is NA
```

<br><br>

## Common Functions & NAs

You don't always have to remove missing data in advance. Many R functions allow you to handle missing values within the
function itself, preventing unnecessary data loss. The best way to check if a function has an option for handling
missing values is to look at its help file using `?function_name`.

<br>

### The na.rm() option in commands.

If there is missing data, then by default, R will set the answer to any statistics to also be missing.

Remember this is our test data

```{r}
print(testdata)
```

Taking the mean of z results in NA

```{r}
# Take the mean of column B
mean(testdata$z)

# Take the correlation between A and B
cor(testdata$x,testdata$z)

```

But many commands include an option (often `na.rm=TRUE`) which you can add to ignore missing data

```{r}
# na.rm  e.g. remove NAs = TRUE
mean(testdata$z, na.rm=TRUE)
sd(testdata$z, na.rm=TRUE)
```

or sometimes I look in the help file (or on google) for the exact terminology that I need. For example, for correlation,
there are several options including

```{r}
#the cor command doesn't follow the pattern.
cor(testdata$x,testdata$z,use = "complete.obs")


```

### Tidyverse & Missing Data {.unnumbered}

Tidyverse generally follows an **explicit missing data handling** approach. Instead of relying on `na.action`, most
functions require you to handle missing values before applying them. For example:

-   **dplyr** functions (`mutate()`, `summarize()`) require `na.rm = TRUE` explicitly.
-   **tidyr** offers `drop_na()` and `replace_na()` for explicit handling.
-   **ggplot2** ignores `NA` by default unless specified.

<br>

### Correlation matrices and missing data {.unnumbered}

Instead of removing entire rows, **pairwise deletion** calculates correlations using all available data for each pair of
variables.

```{r,eval=FALSE}
cor(testdata, use = "pairwise.complete.obs")  
# Correlation matrix with pairwise deletion
```

**When to use this?**

-   Useful when you want to retain as much data as possible.
-   Assumes missingness is random across variables.
-   Can lead to different sample sizes per correlation, which may impact interpretations.

<br>

### Regression Models & missing data {.unnumbered}

\*\*I SUGGEST na.exclude UNLESS YOU HAVE A VERY GOOD REASON\*!\*\*

Many modeling functions in R allow specifying how to handle missing data using `na.action`.

```{r}
# Default: Removes rows with missing values
model1 <- lm(z ~ x, data = testdata, na.action = na.omit)

# Keeps track of missing values and preserves original structure in predictions
model2 <- lm(z ~ x, data = testdata, na.action = na.exclude)
```

<br>

#### Omit vs. Exclude: When to Use Each?

| Method | What It Does | When to Use |
|-----------------------------|---------------------------------------|-----------------------------------------------------|
| `na.omit` | Removes rows with missing data and drops them from predictions | When missing data is truly random and you don’t need predictions for those cases |
| `na.exclude` | Removes rows with missing data but keeps `NA` in predictions | When you need predictions to include `NA` in the same rows as the original missing data |

#### Example: How `na.exclude` Works in Predictions

```{r}
predict(model1)  # Predictions only for complete cases
predict(model2)  # Predictions with missing values retained as NA
```

This keeps the structure of the dataset intact, which is important in time series or longitudinal analyses where the
position of missing values matters.
