---
output:
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: yes  
    toc_float:
      collapsed: no  # Ensure Level 1 & 2 remain expanded
      smooth_scroll: yes
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```


## Introduction

Before analyzing a dataset, it's essential to understand its structure, size, and data types. It is then often useful to understand more about the data inside to help with quality control.


#### Suggested libraries {-}

```{r}
library(dplyr)
library(skimr)
library(GGally)
```

The tutorial is based on a small test dataset with four columns:

- `Make` (categorical)
- `MPG` (numerical)
- `Cylinders` (categorical)
- `Transmission` (categorical)

```{r}
# Create a simple dataset with numerical and categorical variables
test_data <- data.frame(
  Make = factor(c("Toyota", "Ford", "Honda", "Chevy", "Nissan", "BMW")),
  MPG = c(30, 25, 27, 22, 29, 24),
  Cylinders = factor(c(4, 6, 4, 8, 4, 6)),
  Transmission = factor(c("Auto", "Manual", "Auto", "Auto", "Manual", "Manual"))
)
```


## Assessing data structure

Before analyzing a dataset, it's essential to understand its structure, size, and data types. This tutorial will guide you through:

1. Viewing the dataset.
2. Checking its dimensions, column names, and structure.
3. Understanding data types and modifying them if necessary.
4. Preparing for summary statistics.

<br>

### Viewing Data

To explore the dataset interactively, you can:

- **Open it in the RStudio Data Viewer**:  
  Run `View(test_data)` in the console.  
  *Note: `View()` should not be placed inside a code chunk because it only works interactively in RStudio and does not produce console output. It often breaks the knitting process*

- **See a tabular preview in the environment pane**:  
  Click on `test_data` in the environment tab.

- **Print the entire dataset** (not recommended for large datasets):  

```{r}
test_data
```

- **View only the first few rows**:

```{r}
head(test_data)
```

<br>


### Column Names

To see what columns exist:

```{r}
names(test_data)
```

### Number of Rows/Columns

To find out how many rows and columns there are:

```{r}
nrow(test_data)  # Number of rows
ncol(test_data)  # Number of columns
dim(test_data)   # Dimensions (rows, columns)
```

### General Structure

To get a compact summary of the dataset, `str()` provides an overview, including the number of observations, variables, and data types.

```{r}
str(test_data)
```

Another alternative is `glimpse()` (from `dplyr`):

```{r}
glimpse(test_data)
```


<br>

### Data-type/Grouped Data

Each column in a dataset has a specific data type (e.g., numeric, character, factor). "Factor" is the R jargon for variables that have a fixed number of unique values (e.g., categories/groups/families).

We can check the data types using:

```{r}
sapply(test_data, class)
```

<br>

If categorical variables are not stored as factors, we can explicitly define them:

```{r}
test_data$Cylinders <- factor(test_data$Cylinders, levels = c(4, 6, 8))
test_data$Transmission <- factor(test_data$Transmission, levels = c("Auto", "Manual"))
```

This ensures:

- Consistent ordering of categorical levels.
- Avoiding unintended sorting in plots or models.

Similarly, if data "should" be numeric, but isn't for some reason, we can convert it using as.numeric() etc.


<br><br>

## Summary Statistics

Summary statistics are a way to summarize the main characteristics of the data inside a  dataset. They can be used to describe the central tendency, dispersion, and shape of a dataset's distribution.
<br>

### General summary functions

We can use the `summary()` function to get a quick overview of numerical variables.

```{r}
summary(test_data)
```

Alternatively, we can calculate specific summary statistics manually using base R and `dplyr`.

```{r}
# Using base R
mean(test_data$MPG)  # Mean MPG
median(test_data$MPG) # Median MPG
sd(test_data$MPG)     # Standard deviation
range(test_data$MPG)  # Minimum and maximum

# Using dplyr
test_data %>%
  summarise(
    Mean_MPG = mean(MPG),
    Median_MPG = median(MPG),
    SD_MPG = sd(MPG),
    Min_MPG = min(MPG),
    Max_MPG = max(MPG)
  )
```

There are also specific summary commands from a variety of packages including

#### Enhanced Summaries with `skimr`

The `skimr` package provides an easy and readable summary of the dataset.

```{r}
skim(test_data)
```

<br>


#### Pairwise Relationships with `GGally`

The `GGally` package provides useful visual summaries of numerical variables.

```{r}
if (!require(GGally)) install.packages("GGally", dependencies = TRUE)
library(GGally)
# Create pairwise scatter plots
GGally::ggpairs(test_data, columns = c("MPG"))
```


### Frequency tables

We can use `table()` or `dplyr::count()` to summarize categorical variables. e.g. count how many rows there are in different categories.  This is especially useful for categorical data. 

The easiest way to do this is using the table command.  For example to see how many cars have each type of transmission:

```{r}
table(test_data$Transmission)
```
So there are three rows with automatic cars and three with manual, We can also make two way tables

```{r}
table(test_data$Transmission, test_data$Cylinders)
```
So there are 2 cars that are automatic with four cylinders. We can also use the count function in dplyr

```{r}
test_data %>% count(Transmission)
```



```{r, results='hide',eval=FALSE, include=FALSE}
#### Pretty table output
# don't include the install line in your code, run it in the console
# install.package("ztable")

library(ztable)
library(magrittr)
options(ztable.type="html")

mytable <- table(HousesNY$Beds, HousesNY$Baths)

my_ztable =ztable(mytable) 
print(my_ztable,caption="Table 1. Basic Table")
```

<br><br>

### Grouped Summary Statistics

We can also compute statistics such as the mean, maximum, and minimum values for different groups.

#### Tidyverse Approach {-}

Using `group_by()` and `summarise()`, we can compute statistics for each species:


```{r}
test_data %>%
  group_by(Transmission) %>%
  summarise(
    Mean_MPG = mean(MPG),
    Median_MPG = median(MPG),
    Count = n()
  )
```



<br>

#### Base R Approach {-}

I admit, tidyverse is better in this case!

```{r}
aggregate(cbind(MPG, MPG) ~ Transmission, data = test_data, 
          FUN = function(x) c(mean = mean(x, na.rm = TRUE), max = max(x, na.rm = TRUE)))
```





